### 2015-05-12 Dissertation/chapter-graybox.tex
Updates from Gray-Box GECCO paper

\chapter{New Problem Domain: gray-box Optimization}
\label{chap-graybox}
All of the work presented so far has focused on the black-box optimization
domain. These problems are characterized by a complete lack of available
problem specific information. All the algorithm has to perform search
is a ``black-box'' which measures the quality of arbitrary proposed
solutions. Optimizers which are successful in this domain are very general
as there are minimal requirements to apply them to a new problem.

@@

\begin{figure}[t]
  \begin{centering}
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{hc-time}
      \end{centering}
      \caption{Seconds per local optima.}
      \label{fig-hc-time}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{hc-radius}
      \end{centering}
      \caption{Moves in Hamming Ball}
      \label{fig-hc-radius}
    \end{subfigure}
  \end{centering}
  \caption{Comparison of performing hill climbing
           using the black box and gray box domains on Nearest Neighbor NK with $K=5$.
           Figure~\ref{fig-hc-time} shows the time required to bring a random
           solution to a local optima, averaged over 900 restarts across 30 instances.
           Figure~\ref{fig-hc-radius} shows the number of moves which must be
           tested by hill climbing when $N=500$ as $r$ increases.}
  \label{fig-hc-compare}
\end{figure}

@@

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{ConnectedSubgraphs}{$v$, $subset$, $closed$, $open$}
    \State $new\_subset \leftarrow subset \cup \{v\}$
    \State $found \leftarrow [new\_subset]$
    \If{$|new\_subset| \geq r$}
      \Return $found$
    \EndIf
    \State $closed\_here \leftarrow \emptyset$
    \State $open' \leftarrow open \cup adjacent(v)$
    \ForAll{$v' \in open'$}
      \If{$v' \notin closed$}
        \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
        \State $closed \leftarrow closed \cup \{v'\}$
        \State $recurse \leftarrow $\Call{ConnectedSubgraphs}{$v'$, $new\_subset$, $closed$, $open'$}
        \State $found \leftarrow found + recurse$
      \EndIf
    \EndFor
    \State $closed \leftarrow closed - closed\_here$
    \State \Return $found$
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or less which
           contain $subset \cup \{v\}$ and no additional vertices in $closed$.}
  \label{fig-connected-subgraphs}
\end{figure}

The problem of finding which subsets must be checked is therefore equivalent to
finding all induced subgraphs of $r$ or less vertices which are connected. For sparse
graphs this can be efficiently calculated using \Call{ConnectedSubgraphs}{} outlined
in Figure~\ref{fig-connected-subgraphs}. To our knowledge this is a novel algorithm
as previous work did not specify how connected subgraphs were found.

$subset$ and $open$ are both empty for the first top level call of \Call{ConnectedSubgraphs}{}, and
$closed$ contains only $v$.
The result of that call is to find all induced connected subgraphs which contain $v$.
Subsequent top level calls expand $closed$ to also include their $v$.
Each recursive level explores how $new\_subset$ can be expanded by including all $v'$ which are adjacent
to at least one vertex in $new\_subset$. Once all subsets including $new\_subset \cup v'$ have been found,
$v'$ is closed to prevent duplicates. However, when recursion ends $closed$ is returned to its previous state.
The maximum recursive depth for this algorithm is $r$ as each level increases $|subset|$ by 1. Each level
makes at most $|open'|$ recursive calls, with $|open'| \leq |open| + c(k-1)$. Finally there must be a top level
call with each of the $N$ variables. This creates a loose bound of $O(r!(c(k-1))^rN)$, which is $O(N)$ under the
assumption that $r$ and $c$ do not increase with $N$. The bound is loose because some portion of recursions
are prevented by $closed$ and because overlapping adjacency will reduce the rate at which $|open'|$ increases.

As this algorithm finds all connected subsets in $O(N)$ time for a fixed $r$, the number of moves
which must be tested to determine if a solution is an $r$-bit local optima is $O(N)$. This means that
while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black box method would require 85 billion
checks, the gray box method requires only 248,000. Even when allowing connections to be completely random,
which results in $c$ increasing with $N$, the gray box method still only requires approximately 375 million checks,
two and a half orders of magnitude less than assuming a black box.

From these conclusions it is possible to modify the hill climber presented in Section~\ref{sec-whitley-ls} to
efficiently find $r$-bit local optima~\cite{chicano:2014:ball}. The only change is that instead of having a move
and $delta$ for each bit, there must be a move and delta for each connected induced subgraph of $r$ or less bits.
The efficiency analysis is unchanged with the exception that the constant increases exponentially with $r$ as
the number of $delta$ values which must be updated each move increases. Still, with $c$, $r$, and $k$ constant
with respect to $N$, it is possible to find $r$-bit hamming ball local optima in $O(N+I)$ time.

\subsection{Preliminary Empirical Comparison}
Figure~\ref{fig-hc-radius} shows how the number of moves increases as the radius of the hamming ball increases.
Even on this relatively small problem ($N=500$) the black box method quickly becomes intractable. The gray box
method is able to use less moves to find 4-bit local optima than the black box needs to find 2-bit local optima,
and when solving 6-bit local optima the black box method requires 653,635 times as many moves as gray box. However,
gray box still increases exponentially as $r$ increases, limiting how large $r$ can grow before becoming intractable.
This relationship depends heavily on $c$ and how variables are distributed among subfunctions.

\begin{comment}
\section{Memetic Algorithms}
~\cite{chen:2011:memetic}

Local search, even $r$-bit hamming ball search, is not sufficient to robustly find global optimums on
highly multi-modal problems. This limitation was shown in Figure~\ref{fig-evals-to-success} with hill climbing
unable to reach the global optima on almost any large size problem.
\end{comment}

>>>>

\chapter{New Problem Domain: Gray-Box Optimization}
\label{chap-graybox}
All of the work presented so far has focused on the black-box optimization
domain. These problems are characterized by a complete lack of available
problem specific information. All the algorithm has to perform search
is a ``black-box'' which measures the quality of arbitrary proposed
solutions. Optimizers which are successful in this domain are very general
as there are minimal requirements to apply them to a new problem.

@@

@@

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{ConnectedInducedSubgraphs}{}
    \State $closed \leftarrow \emptyset$
    \State $found \leftarrow [~]$
    \ForAll{$v \in V$}
      \State $closed \leftarrow closed \cup \{v\}$
      \State $found \leftarrow found + $\Call{CISG}{$v$, $\emptyset$, $closed$, $\emptyset$}
    \EndFor
    \State \Return $found$
  \EndProcedure
  \Procedure{CISG}{$v$, $subset$, $closed$, $open$}
    \State $subset' \leftarrow subset \cup \{v\}$
    \State $found \leftarrow [subset']$
    \If{$|subset'| \geq r$}
      \Return $found$
    \EndIf
    \State $closed\_here \leftarrow \emptyset$
    \State $open' \leftarrow open \cup adjacent(v)$
    \ForAll{$v' \in open'$ such that $v' \notin closed$}
        \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
        \State $closed \leftarrow closed \cup \{v'\}$
        \State $recurse \leftarrow $\Call{CISG}{$v'$, $subset'$, $closed$, $open'$}
        \State $found \leftarrow found + recurse$
    \EndFor
    \State $closed \leftarrow closed - closed\_here$
    \State \Return $found$
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer.}
  \label{fig-connected-subgraphs}
\end{figure}

In order to determine which moves in the hamming-ball must be evaluated,
we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
\Call{CISG}{} is a recursive helper function which finds all subgraphs
which contain a given $subset$ and a given vertex $v$, while excluding
any other vertices added to $closed$. To find all subgraphs, \Call{CISG}{}
is called once for each vertex in the graph, such that $closed$ contains
all previously searched vertices, and $subset=open=\emptyset$. In the initial
call all desired subgraphs which contain $v$ are found, which is why $v$ remains
in $closed$ to prevent duplicate subgraphs from being returned.

At each recursive level \Call{CISG}{} expands $open$ to include any vertices
adjacent to $v$ in the graph. By construction this means that $open'$ contains
all possible ways of adding a single vertex to the current $subset'$. As each $v'$
is tested it is temporarily added to $closed$ to prevent recursive calls
from adding it again to $subset'$.

When applied to the sparse graphs inherent in the gray-box domain, this algorithm
requires \BigO{r!(ck)^rN} time, which reduces to \BigO{N}. The time spent
in each call is dominated by the loop over $open'$. In the worst case, $open'$
increases in size by the full adjacency of $v$, which is bounded by $ck$.
This creates a worse case complexity for a single top level call of
$\prod_{i}^{r} ick = r!(ck)^r$. This must be called once for each
of the $N$ variables resulting in \BigO{r!(ck)^rN}.

As this algorithm finds all connected subsets in \BigO{N} time for a fixed $r$, the number of moves
which must be tested to determine if a solution is an $r$-bit local optima is \BigO{N}. This means that
while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black-box method would require 85 billion
checks, the gray-box method requires only 248,000. Even when allowing connections to be completely random,
which results in $c$ increasing with $N$, the gray-box method still only requires approximately 375 million checks,
two and a half orders of magnitude less than assuming a black-box.

From these conclusions it is possible to modify the hill climber presented in Section~\ref{sec-whitley-ls} to
efficiently find $r$-bit local optima~\cite{chicano:2014:ball}. The only change is that instead of having a move
and $delta$ for each bit, there must be a move and delta for each connected induced subgraph of $r$ or less bits.
The efficiency analysis is unchanged with the exception that the constant increases exponentially with $r$ as
the number of $delta$ values which must be updated each move increases. Still, with $c$, $r$, and $k$ constant
with respect to $N$, it is possible to find $r$-bit hamming ball local optima in \BigO{N+I} time.

\section{Tournament Uniform Crossover: TUX}
\label{sec-tux}
\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-TUX}{}
    \State Create random $solution$
    \State Hamming-Ball Hill Climb $solution$
    \ForAll{$T_i \in T$}
      \If{$T_i$ is empty}
        \State $T_i \leftarrow solution$
        \State \Return
      \EndIf
      \State Cross $solution$ with $T_i$ to create $2^{i+1}$ offspring
      \State Hamming-Ball Hill Climb each offspring
      \State $solution \leftarrow$ best of offspring, $solution$, and $T_i$
      \State $T_i \leftarrow$ empty
    \EndFor
    \State Add $solution$ to end of $T$
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of TUX optimization. $T$ is an
           ordered list of solutions, each position of which could be empty,
           awaiting a crossover partner.}
  \label{fig-TUX}
\end{figure}

Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optima
on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
like all random restart hill climbers, it relies on random initialization to fall
inside the global optima's basin of attraction.

To remedy this limitation, we set out to develop a minimally complex memetic
algorithm to help increase this probability. Figure~\ref{fig-TUX} presents
the Tournament Uniform Crossover (TUX) algorithm, which combines simplistic
selection with equal probability uniform crossover, the most basic unbiased crossover, to generate starting
solutions likely to be in the global optima's basin of attraction.

Conceptually TUX iteratively builds a structure similar to a single elimination
bracket for solutions. Each ``match'' in the tournament takes in two candidate solutions,
produces offspring via uniform crossover, applies hill climbing to each offspring, with
the ``winner'' being the best of all those solutions. The tournament ``bracket'' is constructed
iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
emerges and solutions only need to be stored until their partner is found.
TUX is fully elitist but does not prematurely converge. This is because search continuously
integrates new randomly generated solutions through other parts of the bracket. Whenever the
top of the current bracket is reached, TUX doubles the size of the virtual tournament.

When crossing solutions at $T_i$, TUX produces $2^{i+1}$ offspring. This relationship
ensures that in total all levels of the tournament, including random initialization,
perform the same number of hill climbing steps. It also shifts the focus of search toward
areas expected to be of higher fitness. The expectation is also that it becomes progressively
harder to improve solutions the higher up the tournament you advance, so more attempts
are necessary to create new useful solutions.

The primary advantages of TUX are that it does not introduce any new parameters (though it still
requires an $r$ for the hill climber) and is relatively simple to implement. Even so
it allows for learning from previous local optima and as Section~\ref{sec-experiments}
will show it is quite effective at optimization.
